The association of smoking status with SARS-CoV-2 infection, hospitalisation and mortality from COVID-19: A living rapid evidence review with Bayesian meta-analyses (`r paste(current_version)`)

### 2021-08-12 reports/manuscript.Rmd
v12 final

**Aims:** To estimate the association of smoking status with rates of i) infection, ii) hospitalisation, iii) disease severity, and iv) mortality from SARS-CoV-2/COVID-19 disease.

@@

**Findings:** `r paste(current_version)` (searches up to `r paste(tail(search_details$date_screened, 1))`) included `r paste(current_total)` studies with `r paste(nrow(running_meta_count))` ‘good’ and ‘fair’ quality studies included in unadjusted meta-analyses. `r paste(nrow(full_smoking_status))` studies (`r paste(round(nrow(full_smoking_status)/current_total*100,1), "%", sep = "")`) reported current, former and never smoking status with the remainder using broader categories. Recorded smoking prevalence among people with COVID-19 was generally lower than national prevalence. Current compared with never smokers were at reduced risk of SARS-CoV-2 infection (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). Data for former smokers were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) but favoured there being no important association (`r paste(round(1-m2_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1). Former compared with never smokers were at increased risk of hospitalisation (RR = `r paste(round(former_hospitalisation_RR_value, 2))`, CrI = `r paste(round(former_hospitalisation_lci, 2), round(former_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(former_hospitalisation_tau, 2))`), greater disease severity (RR = `r paste(round(former_severity_RR_value, 2))`, CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) and mortality (RR = `r paste(round(former_mortality_RR_value, 2))`, CrI = `r paste(round(former_mortality_lci, 2), round(former_mortality_uci, 2), sep = "-")`, τ = `r paste(round(former_mortality_tau, 2))`). Current compared with never smokers were at increased risk of greater disease severity (RR `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`). Data for current smokers on hospitalisation and mortality were inconclusive (RR = `r paste(round(current_hospitalisation_RR_value, 2))`, 95% CrI = `r paste(round(current_hospitalisation_lci, 2), round(current_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(current_hospitalisation_tau, 2))`; RR = `r paste(round(current_mortality_RR_value, 2))`, 95% CrI = `r paste(round(current_mortality_lci, 2), round(current_mortality_uci, 2), sep = "-")`, τ = `r paste(round(current_mortality_tau, 2))`, respectively) but favoured there being no important associations (`r paste(round(1-m3_a_ecdf(1.1),2)*100, "%", sep = "")` and `r paste(round(1-m7_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1, respectively).

**Conclusions:** Compared with never smokers, current smokers appear to be at reduced risk of SARS-CoV-2 infection and increased risk of greater disease severity, while former smokers appear to be at increased risk of hospitalisation, greater disease severity and mortality from COVID-19. However, it is uncertain whether these associations are causal.

@@

There are several reviews that fall within the scope of smoking and COVID-19 [@vardavas_covid-19_2020; @emami_prevalence_2020; @alqahtani_prevalence_2020; @patanavanich_smoking_2020; @berlin_covid-19_2020; @farsalinos_systematic_2020; @grundy_smoking_2020]. We aimed to produce a rapid synthesis of available evidence pertaining to the rates of infection, hospitalisation, disease severity and mortality from SARS-CoV-2/COVID-19 stratified by smoking status. Given the increasing availability of data on this topic, this is a living review with regular updates. As evidence accumulates, the review will be expanded to include studies reporting COVID-19 outcomes by alternative nicotine use (e.g., nicotine replacement therapy or e-cigarettes).

@@

This is a living evidence review  which is updated as new evidence becomes available [@elliott_living_2014]. We adopted recommended best practice for rapid evidence reviews, which involved limiting the search to main databases and having one reviewer extract the data and another verify [@tricco_scoping_2015]. This study was not pre-registered but evolved from a report written for a UK medical society [@simons_smoking_2020]. The most recent version of this living review is available [here](https://www.qeios.com/read/latest-UJR2AW). Version 7 of this living review has been published in a peer-reviewed journal [@simons_association_2020-1]. A completed Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) checklist is included in Supplementary file 1.

@@

  1. Were primary research studies using experimental (e.g. randomised controlled trial), quasi-experimental (e.g. pre- and post-test) or observational (e.g. case-control, retrospective cohort, prospective cohort) study designs;
  2. Included adults aged 16+ years;
  3. Recorded as outcome i) results of a SARS-CoV-2 diagnostic test (including antibody assays), ii) clinical diagnosis of COVID-19, iii) hospitalisation with COVID-19, iv) severity of COVID-19 disease in those hospitalised or v) mortality from COVID-19;
Reported any of the outcomes of interest by self-reported or biochemically verified smoking status (e.g. current smoker, former smoker, never smoker) or current vaping and nicotine replacement therapy (NRT) use;
  4. Were available in English;
  5. Were published in a peer-reviewed journal, as a pre-print or a public health report by reputable bodies (e.g. governments, scientific societies).

@@

A narrative synthesis was conducted. Data from ‘good’ and ‘fair’ quality studies were pooled in R v.3.6.3 [@team_r_2013]. In a living review where new data are regularly added to the analyses, it may be more appropriate to use a Bayesian (as opposed to frequentist) approach where prior knowledge is used in combination with new data to estimate a posterior risk distribution. A Bayesian approach mitigates the issue of performing multiple statistical tests, which can inflate family-wise error. A series of random-effects hierarchical Bayesian meta-analyses were performed with the brms [@burkner_advanced_2017] package to estimate the relative risk for each comparison with accompanying 95% credible intervals (CrIs). We first defined prior distributions for the true pooled effect size (µ) and the between-study heterogeneity (τ), with µ specified as a normal distribution with a mean equal to the derived point estimate from each comparison of interest in the immediately preceding version of this living review, and τ specified as a half-Cauchy distribution with a mean of 0 and standard deviation of 1. The half-Cauchy distribution was selected to reflect prior knowledge that high levels of between-study heterogeneity are more likely than lower levels. Markov Chain Monte Carlo methods (20,000 burn-ins followed by 80,000 iterations) were then used to generate a risk distribution for each study, in addition to a pooled effect for the posterior risk distribution. We report forest plots with the pooled effect for the posterior risk distribution displayed as the median relative risk with an accompanying 95% CrIs. We used the empirical cumulative distribution function (ECDF) to estimate the probability of there being a 10% reduction or 10% increase in relative risk (RR) (i.e. RR ≥1.1 or RR ≤0.9). Due to a lack of indication as to what constitutes a clinically or epidemiologically meaningful effect (e.g. with regards to onward disease transmission or requirements for intensive care beds), we deemed a 10% change in risk as small but important. Where data were inconclusive (as indicated by CrIs crossing RR = 1.0), to disambiguate whether data favoured no effect or there being a small but important association, we estimated whether there was ≥75% probability of RR ≥1.1 or RR ≤0.9.

@@

`r paste(nrow(table_2_word))` studies provided data on SARS-CoV-2 infection for people meeting local testing criteria by smoking status (see [Table 2](#table-2)). Meta-analyses were performed for `r paste(nrow(good_quality_2))` ‘good’ and `r paste(nrow(fair_quality_2))` ‘fair’ quality studies (see [Figure 3](#fig-3) and [4](#fig-4)). Current smokers were at reduced risk of testing positive for SARS-CoV-2 compared with never smokers (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). The six good quality studies each reported point estimates less than 1, although the CrI for two of the six studies crossed 1. The probability of current smokers being at reduced risk of infection compared with never smokers (RR ≤0.9) was >99%. Former compared with never smokers were at increased risk of testing positive, but data were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) and favoured there being no important association. The probability of former smokers being at increased risk of infection (RR ≥1.1) compared with never smokers was < 1%. Results were materially unchanged in sensitivity analyses.

@@

`r paste(nrow(table_4_word))` studies reported disease severity in hospitalised patients stratified by smoking status (see [Table 4](#table-4)). Severe (as opposed to non-severe) disease was broadly defined as requiring intensive treatment unit (ITU) admission, requiring oxygen as a hospital inpatient or in-hospital death. Meta-analyses were performed for 16 ‘fair’ quality studies (see [Figure 7](#fig-7) and [8](#fig-8)). Current (RR = `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`) and former (RR = `r paste(round(former_severity_RR_value, 2))`, 95% CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) compared with never smokers were at increased risk of greater disease severity. The probability of current and former smokers having increased risk of greater disease severity (RR ≥1.1) compared with never smokers was `r paste(round(1-m5_a_ecdf(1.1),2)*100, "%", sep = "")` and `r paste(round(1-m6_a_ecdf(1.1),2)*100, "%", sep = "")`, respectively. Results were materially unchanged in two sensitivity analyses.

@@

This living rapid review found uncertainty in the majority of `r paste(current_total)` studies arising from the recording of smoking status. Notwithstanding this uncertainty, compared with overall adult national prevalence estimates, recorded current smoking rates in most studies were lower than expected. In a subset of good and fair quality studies (n = `r paste(nrow(good_quality_2)+nrow(fair_quality_2))`), current but not former smokers had a reduced risk of testing positive for SARS-CoV-2 but current smokers appeared somewhat more likely to present for testing and/or receive a test. Current compared with never smokers had an increased risk of greater disease severity. Data for current smokers on the risk of hospitalisation and mortality were inconclusive, and favoured there being no important associations. Former smokers were at increased risk of hospitalisation, disease severity and mortality compared with never smokers.

@@

    1.	An estimated 20% (95% CI = 17-25%) of COVID-19 cases are asymptomatic [@buitrago-garcia_occurrence_2020], with some evidence suggesting younger people are more likely to be asymptomatic [@kronbichler_asymptomatic_2020]. Testing is hence likely limited in some subgroups, with the potential for these groups to include an overrepresentation of current smokers.   

@@

This living rapid evidence review was limited by having a single reviewer extracting data with a second independently verifying the data extracted to minimise errors, restricting the search to one electronic database and one pre-print server and by not including at least three large population surveys due to their reliance on self-reported suspected or confirmed SARS-CoV-2 infection (which means they do not meet our eligibility criteria) [@hopkinson_current_2020; @bowyer_geo-social_2020; @jackson_covid-19_2020]. We also did not include a large, UK-based, representative seroprevalence study [@ward_antibody_2020] in our meta-analyses as the odds of testing positive in former smokers was not reported. However, the odds of infection for current smokers (OR = 0.64, 95% CI = 0.58-0.71) was in concordance with the pooled estimate in our meta-analysis. Population surveys – particularly with linked data on confirmed infection or antibodies – will be included in future review versions to help mitigate some of the limitations of healthcare based observational studies. The comparisons of current and former smoking prevalence in the included studies with national prevalence estimates did not adjust observed prevalence for the demographic profile of those tested/admitted to hospital. Other reviews focused on this comparison have applied adjustments for sex and age, and continue to find lower than expected prevalence – notwithstanding the issues complicating interpretation described above [@farsalinos_systematic_2020].

@@

Further scientific research is needed to resolve the mixed findings summarised in our review. First, clinical trials of the posited therapeutic effect of nicotine could have important implications both for smokers and for improved understanding of how the SARS-CoV-2 virus causes disease in humans. Such trials should focus on medicinal nicotine (as smoked tobacco is a dirty delivery mechanism that could mask beneficial effects) and potentially differentiate between different modes of delivery (i.e. inhaled vs. ingested) since this can affect pharmacokinetics [@shahab_novel_2013] and potential therapeutic effects. A second research priority would be a large, representative (randomly sampled) population survey with a validated assessment of smoking status which distinguishes between recent and long-term ex-smokers – ideally biochemically verified – and assesses seroprevalence and links to health records.

@@

Across `r paste(current_total)` studies, recorded current but not past smoking prevalence was generally lower than national prevalence estimates. Current smokers were at reduced risk of testing positive for SARS-CoV-2 and at increased risk of greater disease severity. Former smokers were at increased risk of hospitalisation, disease severity and mortality compared with never smokers.

>>>>

**Aims:** To estimate the association of smoking status with rates of i) infection, ii) hospitalisation, iii) disease severity in hospitalised patients, and iv) mortality from SARS-CoV-2/COVID-19 disease.

@@

**Findings:** `r paste(current_version)` (searches up to `r paste(tail(search_details$date_screened, 1))`) included `r paste(current_total)` studies with `r paste(nrow(running_meta_count))` ‘good’ and ‘fair’ quality studies included in unadjusted meta-analyses. `r paste(nrow(full_smoking_status))` studies (`r paste(round(nrow(full_smoking_status)/current_total*100,1), "%", sep = "")`) reported current, former and never smoking status with the remainder using broader categories. Recorded smoking prevalence among people with COVID-19 was generally lower than national prevalence. Current compared with never smokers were at reduced risk of SARS-CoV-2 infection (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). Data for former smokers were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) but favoured there being no important association (<1% probability of RR ≥1.1). Former compared with never smokers were at increased risk of hospitalisation (RR = `r paste(round(former_hospitalisation_RR_value, 2))`, CrI = `r paste(round(former_hospitalisation_lci, 2), round(former_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(former_hospitalisation_tau, 2))`), greater disease severity (RR = `r paste(round(former_severity_RR_value, 2))`, CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) and mortality (RR = `r paste(round(former_mortality_RR_value, 2))`, CrI = `r paste(round(former_mortality_lci, 2), round(former_mortality_uci, 2), sep = "-")`, τ = `r paste(round(former_mortality_tau, 2))`). Current compared with never smokers were at increased risk of greater disease severity (RR `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`). Data for current smokers on hospitalisation and mortality were inconclusive (RR = `r paste(round(current_hospitalisation_RR_value, 2))`, 95% CrI = `r paste(round(current_hospitalisation_lci, 2), round(current_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(current_hospitalisation_tau, 2))`; RR = `r paste(round(current_mortality_RR_value, 2))`, 95% CrI = `r paste(round(current_mortality_lci, 2), round(current_mortality_uci, 2), sep = "-")`, τ = `r paste(round(current_mortality_tau, 2))`, respectively) but favoured there being no important associations (`r paste(round(1-m3_a_ecdf(1.1),2)*100, "%", sep = "")` and `r paste(round(1-m7_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1, respectively).

**Conclusions:** Compared with never smokers, current smokers appear to be at reduced risk of SARS-CoV-2 infection and increased risk of greater in-hospital disease severity, while former smokers appear to be at increased risk of hospitalisation, greater in-hospital disease severity and mortality from COVID-19. However, it is uncertain whether these associations are causal. This version (v12) will be the last regular update; however, yearly updates may continue as new evidence becomes available.

@@

There are several reviews that fall within the scope of smoking and COVID-19 [@vardavas_covid-19_2020; @emami_prevalence_2020; @alqahtani_prevalence_2020; @patanavanich_smoking_2020; @berlin_covid-19_2020; @farsalinos_systematic_2020; @grundy_smoking_2020]. We aimed to produce a rapid synthesis of available evidence pertaining to the rates of infection, hospitalisation, disease severity and mortality from SARS-CoV-2/COVID-19 stratified by smoking status. Given the increasing availability of data on this topic, this was set up in March 2020 as a living review with regular updates.

@@

This is a living evidence review  which is updated as new evidence becomes available [@elliott_living_2014]. We adopted recommended best practice for rapid evidence reviews, which involved limiting the search to main databases and having one reviewer extract the data and another verify [@tricco_scoping_2015]. This study was not pre-registered but evolved from a report written for a UK medical society [@simons_smoking_2020]. The most recent version of this living review is available [here](https://www.qeios.com/read/latest-UJR2AW). Version 7 of this living review has been published in a peer-reviewed journal [@simons_association_2020-1]. A completed Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) checklist is included in Supplementary file 1. This version (v12) will be the last regular update; however, yearly updates may continue as new, high-quality studies (e.g. large, representative/random population studies or experimental studies) become available.

@@

  1. Were primary research studies using experimental (e.g. randomised controlled trial), quasi-experimental (e.g. pre- and post-test) or observational (e.g. case-control, retrospective cohort, prospective cohort) study designs;
  2. Included adults aged 16+ years;
  3. 3.	Recorded as outcome i) results of a SARS-CoV-2 diagnostic test (including antibody assays) in community or hospitalised populations, ii) clinical diagnosis of COVID-19 in community or hospitalised populations, iii) hospitalisation with COVID-19 in community populations testing positive for SARS-CoV-2, iv) severity of COVID-19 disease in hospitalised populations or v) mortality from COVID-19 in community or hospitalised populations; Reported any of the outcomes of interest by self-reported or biochemically verified smoking status (e.g. current smoker, former smoker, never smoker) or current vaping and nicotine replacement therapy (NRT) use;
  4. Were available in English;
  5. Were published in a peer-reviewed journal, as a pre-print or a public health report by reputable bodies (e.g. governments, scientific societies).

@@

A narrative synthesis was conducted. Data from ‘good’ and ‘fair’ quality studies were pooled in R v.3.6.3 [@team_r_2013]. In a living review where new data are regularly added to the analyses, it may be more appropriate to use a Bayesian (as opposed to frequentist) approach where prior knowledge is used in combination with new data to estimate a posterior risk distribution. A Bayesian approach mitigates the issue of performing multiple statistical tests, which can inflate family-wise error. A series of random-effects hierarchical Bayesian meta-analyses were performed with the brms [@burkner_advanced_2017] package to estimate the relative risk for each comparison with accompanying 95% credible intervals (CrIs). We first defined prior distributions for the true pooled effect size (µ) and the between-study heterogeneity (τ), with µ specified as a normal distribution with a mean equal to the derived point estimate from each comparison of interest in the immediately preceding version of this living review, and τ specified as a half-Cauchy distribution with a mean of 0 and standard deviation of 1. The half-Cauchy distribution was selected to reflect prior knowledge that higher levels of between-study heterogeneity are more likely than lower levels. Markov Chain Monte Carlo methods (20,000 burn-ins followed by 80,000 iterations) were then used to generate a risk distribution for each study, in addition to a pooled effect for the posterior risk distribution. We report forest plots with the pooled effect for the posterior risk distribution displayed as the median relative risk with an accompanying 95% CrIs. We used the empirical cumulative distribution function (ECDF) to estimate the probability of there being a 10% reduction or 10% increase in relative risk (RR) (i.e. RR ≥1.1 or RR ≤0.9). Due to a lack of indication as to what constitutes a clinically or epidemiologically meaningful effect (e.g. with regards to onward disease transmission or requirements for intensive care beds), we deemed a 10% change in risk as small but important. Where data were inconclusive (as indicated by CrIs crossing RR = 1.0), to disambiguate whether data favoured no effect or there being a small but important association, we estimated whether there was ≥75% probability of RR ≥1.1 or RR ≤0.9.

@@

`r paste(nrow(table_2_word))` studies provided data on SARS-CoV-2 infection for people meeting local testing criteria by smoking status (see [Table 2](#table-2)). Meta-analyses were performed for `r paste(nrow(good_quality_2))` ‘good’ and `r paste(nrow(fair_quality_2))` ‘fair’ quality studies (see [Figure 3](#fig-3) and [4](#fig-4)). Current smokers were at reduced risk of testing positive for SARS-CoV-2 compared with never smokers (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). The six good quality studies each reported point estimates less than 1, although the CrI for two of the six studies crossed 1 for current smokers (see starred studies in Figure 3). The probability of current smokers being at reduced risk of infection compared with never smokers (RR ≤0.9) was >99%. Former compared with never smokers were had an equivalent risk of testing positive, with data being inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) ), favouring there being no important association. The probability of former smokers being at increased risk of infection (RR ≥1.1) compared with never smokers was < 1%. Results were materially unchanged in sensitivity analyses.

@@

`r paste(nrow(table_4_word))` studies reported disease severity in hospitalised patients stratified by smoking status (see [Table 4](#table-4)). Severe (as opposed to non-severe) disease was broadly defined as requiring intensive treatment unit (ITU) admission, requiring oxygen as a hospital inpatient or in-hospital death. Meta-analyses were performed for 16 ‘fair’ quality studies (see [Figure 7](#fig-7) and [8](#fig-8)). Current (RR = `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`) and former (RR = `r paste(round(former_severity_RR_value, 2))`, 95% CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) compared with never smokers were at increased risk of greater disease severity. The probability of current and former smokers having increased risk of greater disease severity (RR ≥1.1) compared with never smokers was `r paste(round(1-m5_a_ecdf(1.1),2)*100, "%", sep = "")` and >99%, respectively. Results were materially unchanged in two sensitivity analyses.

@@

This living rapid review found uncertainty in the majority of `r paste(current_total)` studies arising from the recording of smoking status. Notwithstanding this uncertainty, compared with overall adult national prevalence estimates, recorded current smoking rates in most studies were lower than expected. In a subset of good and fair quality studies (n = `r paste(nrow(good_quality_2)+nrow(fair_quality_2))`), current but not former smokers had a reduced risk of testing positive for SARS-CoV-2 but current (and former) smokers appeared somewhat more likely to present for testing and/or receive a test. Current compared with never smokers had an increased risk of greater disease severity in hospitalised patients. Data for current smokers on the risk of hospitalisation and mortality were inconclusive and favoured there being no important associations. Former smokers were at increased risk of hospitalisation, disease severity and mortality compared with never smokers.

@@

    1.	An estimated 20% (95% CI = 17-25%) of COVID-19 cases are asymptomatic [@buitrago-garcia_occurrence_2020], with some evidence suggesting younger people are more likely to be asymptomatic [@kronbichler_asymptomatic_2020]. Testing is hence likely limited in some subgroups, with the potential for these groups to include an overrepresentation of current smokers. The proportion of symptomatic cases may also vary depending on viral strain and vaccination status.      

@@

This living rapid evidence review was limited by having a single reviewer extracting data with a second independently verifying the data extracted to minimise errors, restricting the search to one electronic database and one pre-print server and by not including at least three large population surveys due to their reliance on self-reported suspected or confirmed SARS-CoV-2 infection (which means they do not meet our eligibility criteria) [@hopkinson_current_2020; @bowyer_geo-social_2020; @jackson_covid-19_2020]. We also did not include a large, UK-based, representative seroprevalence study [@ward_antibody_2020] in our meta-analyses as the odds of testing positive in former smokers was not reported. However, the odds of infection for current smokers (OR = 0.64, 95% CI = 0.58-0.71) was in concordance with the pooled estimate in our meta-analysis. Population surveys – particularly with linked data on confirmed infection or antibodies – should be included in future reviews to help mitigate some of the limitations of healthcare based observational studies. The comparisons of current and former smoking prevalence in the included studies with national prevalence estimates did not adjust observed prevalence for the demographic profile of those tested/admitted to hospital. Other reviews focused on this comparison have applied adjustments for sex and age, and continue to find lower than expected prevalence – notwithstanding the issues complicating interpretation described above [@farsalinos_systematic_2020].

@@

Further scientific research is needed to resolve the mixed findings summarised in our review. First, clinical trials of the posited therapeutic effect of nicotine could have important implications both for smokers and for improved understanding of how the SARS-CoV-2 virus causes disease in humans. Such trials should focus on medicinal nicotine (as smoked tobacco is a dirty delivery mechanism that could mask beneficial effects) and potentially differentiate between different modes of delivery (i.e. inhaled vs. ingested) since this can affect pharmacokinetics [@shahab_novel_2013] and potential therapeutic effects. Of note, a recent study of rhesus macaques reported that the nAChR agonist varenicline has the potential to interact with and inhibit SARS-CoV-2 infection and replication. A second research priority would be a large, representative (randomly sampled) population survey with a validated assessment of smoking status which distinguishes between recent and long-term ex-smokers – ideally biochemically verified – and assesses seroprevalence and links to health records.

@@

Across `r paste(current_total)` studies, recorded current but not past smoking prevalence was generally lower than national prevalence estimates. Current smokers were at reduced risk of testing positive for SARS-CoV-2 and at increased risk of greater in-hospital disease severity. Former smokers were at increased risk of hospitalisation, in-hospital disease severity and mortality compared with never smokers.
